Simple connectome inference from partial correlation statistics in calcium imaging

### 2014-06-19 article.tex
typos

The human brain is a complex biological organ made of 100 billions of neurons,
each connected to 7000 other neurons on average. Unfortunately, direct
observations of the connectome, the wiring diagram of the brain, is not yet
technically feasible. Without being perfect, calcium imaging currently allows
the realtime and simultaneous observation of neuron activities from thousands
of neurons, producing individual time series representing their fluoresence
intensity. From these data, the connectome inference problem consists in
retrieving the synaptic connections between neurons on the basis of the
fluoresence time series. In particular, this problem is often made difficult
because of experimental issues, including masking effects (making some of the
neurons not to be observed or confounded with others), the low sampling rate of
the optical device with respect to the neural activity speed or the slow decay
of fluoresence.

@@

In this paper, we present a simplified -- yet nearly as good -- version of the
winning method\footnote{Code available at \url{https://github.com/asutera
/kaggle-connectomics}} of the Connectomics
Challenge\footnote{\url{http://connectomics.chalearn.org}}, as a simple and
theoretically grounded approach based on signal processing techniques and
partial correlation statistics. The paper is structured as follows: Section
\ref{sec:filter} describes the signal processing methods applied on fluorescent
calcium time series; Section \ref{sec:inference} then presents the proposed
approach and its theoretical properties; Section \ref{sec:results} provides an
empirical analysis and comparison with other network inference methods, while
Section \ref{sec:conclusion} finally discusses our work and provides further
research directions. Additionnaly, Appendix \ref{app:optimized} further
describes, in all details, our actual winning method, giving slightly better
results than the method presented in the paper but highly tuned for the
challenge.

@@

Under the simplifying assumption that neurons are on-off units, characterized
by short periods of intense activities, or peaks, and longer periods of
inactivity, the first part of our algorithm consists in cleaning the raw
fluoresence data. More specifically, time series are processed using standard
signal processing filters in order to remove noise due light scattering
effects, account for fluorescence  low decay and reduce the importance of
global high activity in the network. The overall process is illustrated on
Figure~\ref{fig:filtered-signal}.

@@

Under that assumption that random variables follow a joint Gaussian
distribution $\mathcal{N}(\mu; \Sigma)$ of mean vector $\mu$ and covariance
matrix $\Sigma$, $X_i$ and $X_j$ are conditionnally independent with respect to
the other variables $X^{-i,j}$ if their \textit{partial correlation} statistic
is null~\citep{koller2009probabilistic}, i.e., if
\begin{equation}
\rho_{X_i, X_j | X^{-i,j}} = -\frac{\Sigma^{-1}_{ij}}{\sqrt{\Sigma^{-1}_{ii} \Sigma^{-1}_{jj}}} = 0.
\end{equation}
Conversely, $X_i$ and $X_j$ are conditionally dependent if $\rho_{X_i, X_j | X^{-i,j}} \neq 0$,
where $\rho_{X_i, X_j | X^{-i,j}}$ can now be interpreted as the degree of association
between the two variables (i.e., the larger the value, the stronger the interaction).
Together, the partial correlation statistics correspond to the inverse
$\Sigma^{-1}$ of the covariance matrix, also known as the concentration of
precision matrix. Computationnally, they can be estimated efficiently in
$O(M^3)$ (instead of $O(p^3)$) from the $M$ first principal components
extracted through PCA~\citep{bishop2006pattern} on the processed data. In our experiments, $M$ is set
to $0.8 p$, thereby allowing us to further filter noise in the data.

@@

As it can also be seen on Figure \ref{fig:curves}, the GENIE3 performance
increases with deeper trees in addition to be computionnally feasible. The
reason is twofold. First, the deeper the tree, the larger is the conditioning.
As it was observed in the comparison between correlation and partial
correlation, it screens indirect links in the infered network. Second, the
number of small trees need to be much higher than deep trees in order to
observe - and evaluate - all pairs.  However, as suggested by
\cite{louppe2014understanding}, limiting the maximal depth of trees may help
to reduce the misestimation bias in variable importance comuting, i.e.
infering the strengh of a link. Therefore a trade-off of the maximal depth may give the best of both worlds.

@@

\begin{figure}[bth]
\centering
\includegraphics[width=\linewidth]{images/curves}
\caption{Method precision-recall and roc curves. \textcolor{red}{GENI3?}
         $\tau=0.11$}
\label{fig:curves}
\end{figure}

@@

The area under the ROC curve is not consistent metric
with respect to the problem statement. A clear hierachy is also difficult to derive based on ROC curve of Figure \ref{fig:curves}. We advice to avoid the AUROC
for the assessment of network inference method. An alternative metric would
be the area under the precision-recall curve (AUPRC). Those
conclusions are in line with \cite{schrynemackers2013protocols}.

@@

In this paper, we present the winning method of the Connectomics challenge.
This approach is simple and theorically grounded. We also evaluate the method
in a more detailed way and perform a comparison with other inference methods.
We also discuss the relevance of the metric used in the challenge.

In the discussion, we give hints about limiting the depth of trees in the GENIE3
method to infer the network. It should be interesting to see if there is a
optimum value for the maxmimal depth. And of course, it may also rewarding to increase considerably the number of trees to get more stable predictions.

In previous study \citep{shipley2002cause}, it has been suggested to
consider correlation coefficients of all possible orders,
i.e. conditioned on every subsets of the set of $p-2$ other variables. This
would take into account multiple indirect paths between a given pair of
variables. Instead of considering only the highest order, it may be useful to
study with accuracy the infered network at all orders. In this case, it may not
be very relevant but interesting in case of dymanical networks.

As we mentionned, in order to maximize the ROC AUC, it was not necessary to
discover causal relationships. It may be, in future works, promising to focus
on that particular topic.

\section*{Acknowledgements}
A. Joly and G. Louppe are research fellows of
the FNRS, Belgium.  A. Sutera is recipient of
a F.R.I.A. fellowship of F.R.S.-FNRS, Belgium.
This work is supported by PASCAL2 and the IUAP DYSCO, initiated by the
Belgian State, Science Policy Office.

>>>>

The human brain is a complex biological organ made of 100 billions of neurons,
each connected to 7000 other neurons on average. Unfortunately, direct
observations of the connectome, the wiring diagram of the brain, is not yet
technically feasible. Without being perfect, calcium imaging currently allows
the real-time and simultaneous observation of neuron activities from thousands
of neurons, producing individual time series representing their fluorescence
intensity. From these data, the connectome inference problem consists in
retrieving the synaptic connections between neurons on the basis of the
fluorescence time series. In particular, this problem is often made difficult
because of experimental issues, including masking effects (making some of the
neurons not to be observed or confounded with others), the low sampling rate of
the optical device with respect to the neural activity speed or the slow decay
of fluorescence.

@@

In this paper, we present a simplified -- yet nearly as good -- version of the
winning method\footnote{Code available at \url{https://github.com/asutera
/kaggle-connectomics}} of the Connectomics
Challenge\footnote{\url{http://connectomics.chalearn.org}}, as a simple and
theoretically grounded approach based on signal processing techniques and
partial correlation statistics. The paper is structured as follows: Section
\ref{sec:filter} describes the signal processing methods applied on fluorescent
calcium time series; Section \ref{sec:inference} then presents the proposed
approach and its theoretical properties; Section \ref{sec:results} provides an
empirical analysis and comparison with other network inference methods, while
Section \ref{sec:conclusion} finally discusses our work and provides further
research directions. Additionally, Appendix \ref{app:optimized} further
describes, in all details, our actual winning method, giving slightly better
results than the method presented in the paper but highly tuned for the
challenge.

@@

Under the simplifying assumption that neurons are on-off units, characterized
by short periods of intense activities, or peaks, and longer periods of
inactivity, the first part of our algorithm consists in cleaning the raw
fluorescence data. More specifically, time series are processed using standard
signal processing filters in order to remove noise due light scattering
effects, account for fluorescence  low decay and reduce the importance of
global high activity in the network. The overall process is illustrated on
Figure~\ref{fig:filtered-signal}.

@@

Under that assumption that random variables follow a joint Gaussian
distribution $\mathcal{N}(\mu; \Sigma)$ of mean vector $\mu$ and covariance
matrix $\Sigma$, $X_i$ and $X_j$ are conditionally independent with respect to
the other variables $X^{-i,j}$ if their \textit{partial correlation} statistic
is null~\citep{koller2009probabilistic}, i.e., if
\begin{equation}
\rho_{X_i, X_j | X^{-i,j}} = -\frac{\Sigma^{-1}_{ij}}{\sqrt{\Sigma^{-1}_{ii} \Sigma^{-1}_{jj}}} = 0.
\end{equation}
Conversely, $X_i$ and $X_j$ are conditionally dependent if $\rho_{X_i, X_j | X^{-i,j}} \neq 0$,
where $\rho_{X_i, X_j | X^{-i,j}}$ can now be interpreted as the degree of association
between the two variables (i.e., the larger the value, the stronger the interaction).
Together, the partial correlation statistics correspond to the inverse
$\Sigma^{-1}$ of the covariance matrix, also known as the concentration of
precision matrix. Computationally, they can be estimated efficiently in
$O(M^3)$ (instead of $O(p^3)$) from the $M$ first principal components
extracted through PCA~\citep{bishop2006pattern} on the processed data. In our experiments, $M$ is set
to $0.8 p$, thereby allowing us to further filter noise in the data.

@@

As it can also be seen on Figure \ref{fig:curves}, the GENIE3 performance
increases with deeper trees in addition to be computationally feasible. The
reason is twofold. First, the deeper the tree, the larger is the conditioning.
As it was observed in the comparison between correlation and partial
correlation, it screens indirect links in the inferred network. Second, the
number of small trees need to be much higher than deep trees in order to
observe - and evaluate - all pairs.  However, as suggested by
\cite{louppe2014understanding}, limiting the maximal depth of trees may help
to reduce the misestimation bias in variable importance computing, i.e.
inferring the strength of a link. Therefore a trade-off of the maximal depth may give the best of both worlds.

@@

\begin{figure}[bth]
\centering
\includegraphics[width=\linewidth]{images/curves}
\caption{Method precision-recall and ROC curves. \textcolor{red}{GENI3?}
         $\tau=0.11$}
\label{fig:curves}
\end{figure}

@@

The area under the ROC curve is not consistent metric
with respect to the problem statement. A clear hierarchy is also difficult to derive based on ROC curve of Figure \ref{fig:curves}. We advice to avoid the AUROC
for the assessment of network inference method. An alternative metric would
be the area under the precision-recall curve (AUPRC). Those
conclusions are in line with \cite{schrynemackers2013protocols}.

@@

In this paper, we present the winning method of the Connectomics challenge.
This approach is simple and theoretically grounded. We also evaluate the method
in a more detailed way and perform a comparison with other inference methods.
We also discuss the relevance of the metric used in the challenge.

In the discussion, we give hints about limiting the depth of trees in the GENIE3
method to infer the network. It should be interesting to see if there is a
optimum value for the maximal depth. And of course, it may also rewarding to increase considerably the number of trees to get more stable predictions.

In previous study \citep{shipley2002cause}, it has been suggested to
consider correlation coefficients of all possible orders,
i.e. conditioned on every subsets of the set of $p-2$ other variables. This
would take into account multiple indirect paths between a given pair of
variables. Instead of considering only the highest order, it may be useful to
study with accuracy the inferred network at all orders. In this case, it may not
be very relevant but interesting in case of dynamical networks.

As we mentioned, in order to maximize the ROC AUC, it was not necessary to
discover causal relationships. It may be, in future works, promising to focus
on that particular topic.

\section*{Acknowledgments}
A. Joly and G. Louppe are research fellows of
the FNRS, Belgium.  A. Sutera is recipient of
a F.R.I.A. fellowship of F.R.S.-FNRS, Belgium.
This work is supported by PASCAL2 and the IUAP DYSCO, initiated by the
Belgian State, Science Policy Office.
